### 2017-06-16 manuscript/manuscript.Rmd
Clarify writing of part I of tutorial

Fortunately for the empirical sciences, challenges related to organizing and curating materials ("digital assets") across time, space, and personnel have been solved to a high standard in computer science with Version Control Systems (VCS). In the remainder of this article, we introduce a popular VCS called Git, and illustrate its use in the scientific workflow with a hypothetical example project. In the tutorial below, we show how to use the Git VCS by using text commands from the computer's command line. After introducing the fundamentals of Git's VCS functions from the command line, we also show how to use Git easily with a Graphical User Interface (GUI) implemented in the popular R Studio Integrated Development Environment (IDE).

@@

The first operating principle of Git is that your work is organized into independent projects, which Git calls *repositories*. At its core, a repository is a folder on your computer, which is version controlled by Git (you can tell if a folder is "monitored" by git by checking if the folder, or any of its parent folders, contains a hidden `.git` directory). Everything that happens inside a repository is tracked by Git, but the user has full control of what is tracked (everything, by default) and when. Because the user(s) has full control of what and when is tracked, there is a small set of operations the users need to execute every time they wish to log something in Git.

Briefly, when you work in a Git repository, Git stores the state of each file that it monitors, and when any of these files change, Git indicates that they differ from the previously logged state. If the user then is happy with the current changes, she can **add** the changed files to Git's "staging area". If the user then is certain that the files added to the staging area are in good shape, they can **commit** the changes. These two operations are the backbone of using Git to store the state of the project whenever meaningful changes are made.

@@

Most importantly, each state in Git's log contains the full state of the files at that point in time. Users can always go back to an earlier version by "checking out" a previous state of Git's log. That's why these programs are called Version Control Systems.

To establish the operating principles of Git in practice, we now turn to a practical example using a hypothetical project that, we feel, reflects the components of a typical project in experimental psychology. To most clearly show the use of Git, we begin with an empty project with no components.

@@

Implementing reproducibility into the scientific workflow is less time-consuming if it is planned from the onset of the project, rather than attempting to add reproducibility to the project after it has been completed. It is therefore important to organize the project keeping a few key goals in mind (here, we broadly follow established guidelines such as Project TIER recommendations): The files and folders should have easy to understand names (avoid idiosynchratic naming schemes), and the names should indicate the purposes of the files and folders.

The first step is to create a home folder for the project. This folder should have an immediately recognizable name, and should be placed somewhere on your computer where you can find it. We call the example project `git-example`. All the materials related to this project will be placed in subfolders of the home directory, but for now `git-example` is just an empty folder waiting to be filled with data, code, and documents.

@@

Next, you need to navigate to the folder and initialize it as a Git repository. Here, the `git-example` project is in the users `Documents` folder, and we can use `cd ..` in the OS X terminal to navigate up in the folder hierarchy, and `cd <folder>` to navigate into `<folder>`. So assuming that the Terminal opens up in `Documents`, we navigate to `git-example` with (to execute code in Terminal, enter the text into the command line and hit Return):

```bash
cd git-example
```

Then, to turn this folder into a Git repository, execute

```bash
git init
```

This command initializes the folder as a Git repository, and the only change so far has been the addition of a hidden `.git` folder inside `git-example` (and a `.gitattributes` file; but users can ignore these hidden files and folders).

@@

Every project, and therefore every repository, should contain a brief not that explains what the project is about and who to contact about it. This note is usually called a readme, and therefore our first contribution to this project will be a README file. The README file should be a plain text file (i.e. not created with Microsoft Word) that can be read with a simple text editor. This file can now be seen in the `git-example` folder by using the system file viewer. Because we added this file to a Git repository, Git is also aware of it. To see what files have changed since the last status change in the repository (there clearly has been only this one), you can ask for Git's **status**:

```bash
git status
```

Which in this case would return

```bash
Matti [~/Documents/git-example]$ git status
On branch master
Initial commit
Untracked files:
  (use "git add <file>..." to include in what will be committed)
	README
nothing added to commit but untracked files present (use "git add" to track)
Matti [~/Documents/git-example]$ 
```

The relevant output returned from executing this command is the "Untracked files:" part. There, Git tells the user that there is an untracked file (README) in the repository. To keep track of the status of this file, we **add** it to Git by using the command `git add` followed by either a `.` (for adding all untracked files) or `README` for only tracking the `README` file.

```bash
git add README
```

We've now added this file to the staging area, and if we are happy with changes to the file's status (it has been created), we can **commit** the file to Git's versioning system. Commits are the most important Git operation: They signify any meaningful change to the repository, and can be later browsed and compared to one another. As such, it is helpful to attach a small message to each commit, describing why that commit was made. Here, we've created a README file, and our commit command would look as follows:

```bash
git commit -m "Add README file."
```

The quoted text after the `-m` flag is the "commit message". Entering this command to the command line returns a brief description of the commit, such as how many files changed, and how many characters inside those files were inserted and deleted.

@@

Most importantly, the `git-example` project now keeps track of all and any changes to README. To illustrate, we now add some text to README to describe the project (title, what the project is about, who are involved, who to contact), save the file, and then ask for Git's status with `git status` on the command line:

```bash
Matti [~/Documents/git-example]$ git status
On branch master
Changes not staged for commit:
  (use "git add <file>..." to update what will be committed)
  (use "git checkout -- <file>..." to discard changes in working directory)

	modified:   README

no changes added to commit (use "git add" and/or "git commit -a")
```

Git can now tell that the README file has changed, and we can repeat the add and commit steps to permanently record the current state of the project to Git's log:

```bash
git add .  # We use the '.' shortcut 
git commit -m "Populate README with project description."
```

@@

The real importance of these somewhat abstract steps becomes apparent when we consider the Git **log**. We can call

```bash
git log
```

To reveal the commit log of this repository. The main things to notice in the output of this command are that each commit is associated with a unique hash code (long alphanumeric string), which we can use to call for further information (see below); an author (this is where the earlier setup is apparent); a date and time; and the short commit message. Until now, the `git-example` log looks like this:

```bash
Matti [~/Documents/git-example]$ git log
commit 60cbe5c9b4a78e500314f791080381030577a035
Author: Matti Vuorre <mv2521@columbia.edu>
Date:   Tue Jun 13 17:20:27 2017 -0400

    Populate README with project description.

commit 16c475023ecbc99446164187eeaaab10647ac550
Author: Matti Vuorre <mv2521@columbia.edu>
Date:   Tue Jun 13 17:14:14 2017 -0400

    Add README file.
```

To see what exactly changed in the last commit (the log has latest commits at the top), we can call `git show` with the commit's hash (only relevant parts of output shown below):

```bash
Matti [~/Documents/git-example]$ git show 60cbe5c9b4a78e500314f791080381030577a035
commit 60cbe5c9b4a78e500314f791080381030577a035
Author: Matti Vuorre <mv2521@columbia.edu>
Date:   Tue Jun 13 17:20:27 2017 -0400

    Populate README with project description.

diff --git a/README b/README
--- a/README
+++ b/README
@@ -0,0 +1,8 @@
+# Example Git Project
+This example project illustrates the use of Git.
+authors:
+Matti Vuorre <mv2521@columbia.edu>
+James Curley
+2017
```

This output can be investigated for a detailed log of all changes created by that commit. From top, it lists the author of the commit, the commit message, and then the commit's "**diff**" (i.e. what differs in the new version vs the old version of the file.) The current diff shows that 1 file received eight additional lines of text (the part wrapped in @ symbols), and then the additions themselves (the lines prepended with +s).

Although we now understand the fundamentals of using Git to track the states of (and therefore changes to) a repository, this contrived and overly simplistic example doesn't allow full appreciation of the benefits of using Git for version control. To better illustrate Git's functioning, we now fast-forward in the hypothetical example project to a stage in the project where more files and materials have been created.

## (Slightly more) advanced Git

The state of the project now looks like this:

```bash
git-example/
    admin/
        ethics-info.pdf
    experiment-1/
        analysis/
            plan_n.R
    manuscript/
    README
```

Running `git status` now tells that there are two new files (empty folders are ignored). That is, since the previous commit, two files have been added to the project; one a pdf with some administrative information (`ethics-info.pdf`), the other one is an R script for a pre-planned power analysis (`plan_n.R`). We would certainly like to track any changes to the power analysis script, but the ethics information file isn't really something that we need to keep track of---we can't make changes to it anyway. It would be laborious to constantly keep ignoring it when committing changes to Git (especially if we'd like to commit multiple files simultaneously with `git add .`) Git has an elegant solution to specifying which files to keep track of. Because by default all files are tracked, Git uses a special file for instructing which files are to be *ignored* from tracking.

@@

To make Git ignore files, we simply add a plain text file called `.gitignore` to the home folder of the repository. Each row of this file should specify a file or a folder (or a regular expression) that Git should ignore. In our example we want to make Git ignore the `admin/` folder entirely, and any file with the .pdf extension inside `manuscript/. The example .gitignore file looks like this (lines beginning with `#` are comments):

@@

# Ignore all .pdf files in the 'manuscript' folder
manuscript/*.pdf
```

@@

Because there are now two untracked files, and we would like to keep a clean history of what has happened in the project, we create two separate commits: One for the .gitignore file, and one for the power analysis file.

@@

After this last commit, if we ever want to know in the future what our planned sample size was, we can come back to this commit with `git log` or `git show` and see what was inside the newly created power analysis file. For instance, if new information suggests that we should change assumed effect size in the power analysis, we can simply edit and save the file, then add and commit the changes to Git with a helpful message that logs this important event in the Git memory of the repository.

This possibility of "rewinding history" is especially useful for files that might go under multiple revisions, or if we might be interested when and in what order the files were created. One might consider committing a power analysis as a small personal pre-registration of a part of the research plan.

@@

[^git-version]: As of the writing of this article, the current version of Git is 2.13.1.
[^git-name]: The creator of Git, Linus Torvalds, named Git after himself as "the stupid content tracker" (@mcmillan_after_2005; <https://git-scm.com/docs/git.html>)
[^simple]: By "simple", we only mean that the technical aspects of the project, related to curating materials related to it, are simple. We of course do not suggest that there might be anything theoretically or scientifically simple in such projects.
[^vcs-collaborate]: The software we present below is used used by major software developers such as Microsoft, Google, and Facebook on code bases with hundreds of contributors.
[^cli-links]: If using a text-based command line seems challenging, Codecademy (<https://www.codecademy.com/learn/learn-the-command-line>) has a free interactive online tutorial, and MIT offers a free online game to teach using the command line (<http://web.mit.edu/mprat/Public/web/Terminus/Web/main.html>).

>>>>

Fortunately for the empirical sciences, challenges related to organizing and curating materials ("digital assets") across time, space, and personnel have been solved to a high standard in computer science with Version Control Systems (VCS). In the remainder of this article, we introduce a popular VCS called Git, and illustrate its use in the scientific workflow with a hypothetical example project. In the tutorial below, we show how to use the Git VCS by using text commands from the computer's command line. After introducing the fundamentals of Git's VCS functions from the command line, we also show how to use Git easily with a Graphical User Interface (GUI) implemented in the popular R Studio Integrated Development Environment (IDE; @rstudio_team_rstudio:_2016).

@@

The first operating principle of Git is that your work is organized into independent projects, which Git calls *repositories*[^git-link-submodules]. A repository is a folder on your computer which is version controlled by Git (you can tell if a folder is "monitored" by git by checking if the folder, or any of its parent folders, contains a hidden `.git` directory). Everything that happens inside a repository is tracked by Git, but you have full control of what is committed to Git's history and when. Because you have full control of what and when is committed to history, there is a small set of operations you need to know[^git-cheat-sheet].

Briefly, when you work in a Git repository (make changes to files within it), Git monitors the state of each file, and when they change Git knows that they differ from the previously logged state. If you are happy with the current changes, you **add** the changed files to Git's "staging area". If you then are certain that the changes in the staging area are desirable, you **commit** the changes. These two operations are the backbone of using Git to store the state of the project whenever meaningful changes are made. Importantly, each commit in the repository's history contains information to recover the full state of the project at that point in time. Users can always go back to an earlier version by **checking out** a previous state from Git's history. This core of the Git workflow is illustrate in Figure \@ref(fig:git-diagram).

@@

To understand the Git workflow in practice, we now turn to a practical example using a hypothetical project. Git can be added to a project at any stage of the project's life cycle, but to most clearly show its use, we begin with an empty project with nothing in it.

@@

Implementing reproducibility into the scientific workflow is less time-consuming and effortful if it is planned from the onset of the project, rather than added to the project after all the work has been completed. It is therefore important to organize the project keeping a few key goals in mind (here, we follow guidelines such as the Project TIER recommendations[^tier-link]): The files and folders should have easy to understand names (avoid idiosynchratic naming schemes), and the names should indicate the purposes of the files and folders.

The first step is to create a home folder for the project. This folder should have an immediately recognizable name, and should be placed somewhere on your computer where you can find it. We call the example project `git-example`. All the materials related to this project will be placed in subfolders of the home directory, but for now `git-example` is just an empty folder waiting to be filled with data, code, and documents. To make the example concrete, you can follow the tutorial by typing out the following commands on your own computer.

@@

Next, you need to create a new folder for the Git repository. First, choose an appropriate folder on your computer (such as `User/Documents/`) where you'd like to create the project. You can either use the system's file navigator (Finder / File Explorer) to create this folder, or the following commands in the command line: Navigate to the desired folder by using `cd Documents` to move into the `Documents` folder (assuming it exists in the folder where you currently are), and `cd ..` to move out of the folder (to its containing folder.) Once you are in the folder where you want to create the project, type `mkdir git-example` to make the `git-example` directory, then `cd git-example` to move into it. Once you are in the project's home folder (you can verify where you are by typing `pwd`), you want to turn the folder into a project by initializing Git:

```bash
> git init
```

This command initializes the folder as a Git repository, and the only change so far has been the addition of a hidden `.git` folder inside `git-example` (and a `.gitattributes` file. Users can ignore these hidden files and folders, however they can be shown by typing `ls -la`.) Now that the folder is initialized as a Git repository, Git monitors any changes within it, and allows you to add and commit these changes (Figure \@ref(fig:git-diagram).)

@@

Every project (repository) should contain a brief note explaining what the project is about and who to contact. This note is usually called a readme, and therefore our first contribution to this project will be a README file (this file is so important that it has become standard practice to write it in capital letters). The README file should be a plain text file (i.e. not created with Microsoft Word) that can be read with a simple text editor. You can create this file with any text editor (Macs have TextEdit app installed by default, which allows creating plain text files). We created the file with some text in it, and it can now be seen in the `git-example` folder, either by using the system file viewer or `ls` in the command line. Because we added this file to a Git repository, Git is also aware of it. To see what files have changed since the last status change in the repository (there clearly has been only this one), you can ask for Git's **status**:

```bash
> git status

On branch master
Initial commit
Untracked files:
  (use "git add <file>..." to include in what will be committed)
	README
nothing added to commit but untracked files present (use "git add" to track)
```

The relevant output returned from executing this command is the "Untracked files:" part. There, Git tells that there is an untracked file (README) in the repository. To start tracking changes in this file, we **add** it to Git by using the command `git add` followed by either a `.` (for adding all untracked files) or `README` for only tracking the `README` file.

```bash
> git add README
```

We've now added this file to the staging area (Figure \@ref(fig:git-diagram)), and if we are happy with changes to the file's status, we can **commit** the file to Git's history. Commits are essential Git operations: They signify meaningful changes to the repository, and can be later browsed and compared to one another. As such, it is helpful to attach a small message to each commit, describing why that commit was made. Here, we've created a README file, and our commit command would look as follows:

```bash
> git commit -m "Add README file."
```

The quoted text after the `-m` argument is the *commit message*. Entering this command to the command line returns a brief description of the commit, such as how many files changed, and how many characters inside those files were inserted and deleted.

@@

The `git-example` project (or rather, Git) now keeps track of all and any changes to README. To illustrate, you can change the text in the README file with a text editor, save the file, and then ask for Git's status with `git status` on the command line:

```bash
> git status

On branch master
Changes not staged for commit:
  (use "git add <file>..." to update what will be committed)
  (use "git checkout -- <file>..." to discard changes in working directory)
	modified:   README
no changes added to commit (use "git add" and/or "git commit -a")
```

Git can tell that the README file has changed. It is often useful to know exactly *how* a file has changed, before committing it. To view **diff**erences to a file not yet committed, use `git diff <file>`. It shows changes within `<file>`, line by line, highlighting removals with red and added lines with green. Once you are happy with the changes, you can repeat the add and commit steps from above to permanently record the current state of the project to Git's log (below we use the `.` shortcut for adding all changes):

```bash
> git add .
> git commit -m "Populate README with project description."
```

@@

The real importance of these somewhat abstract steps becomes apparent when we consider the Git **log**. To reveal the commit log of your repository, call:

```bash
> git log
```

The main outputs of this command show that each commit is identified with a unique hash code (long alphanumeric string), which we can use to call for further information (see below); an author; a date and time; and the short commit message. Executing `git log` on our example project at this stage returns this:

```bash
commit 60cbe5c9b4a78e500314f791080381030577a035
Author: Matti Vuorre <mv2521@columbia.edu>
Date:   Tue Jun 13 17:20:27 2017 -0400
    Populate README with project description.

commit 16c475023ecbc99446164187eeaaab10647ac550
Author: Matti Vuorre <mv2521@columbia.edu>
Date:   Tue Jun 13 17:14:14 2017 -0400
    Add README file.
```

To see what exactly changed in the last commit (the log has latest commits at the top), you can call `git show` with the commit's hash (only relevant parts of output shown below):

```bash
> git show 60cbe5c9b4a78e500314f791080381030577a035
commit 60cbe5c9b4a78e500314f791080381030577a035
Author: Matti Vuorre <mv2521@columbia.edu>
Date:   Tue Jun 13 17:20:27 2017 -0400
    Populate README with project description.
diff --git a/README b/README
--- a/README
+++ b/README
@@ -0,0 +1,8 @@
+# Example Git Project
+This example project illustrates the use of Git.
+authors:
+Matti Vuorre <mv2521@columbia.edu>
+James Curley
+2017
```

This output is a detailed log of all changes in that commit. From top, it lists the author, the message, and then the commit's "**diff**" (i.e. what differs in the new version vs the old version of the file.) The current diff shows that 1 file received eight additional lines of text (the part wrapped in @ symbols), and then the additions themselves (the lines prepended with +s).

Although you have now seen the fundamentals of using Git to track the states of (and therefore changes to) a repository, this contrived and overly simplistic example doesn't allow full appreciation of the benefits of using Git for version control. To better illustrate Git's functioning, we now fast-forward in the hypothetical example project to a stage in the project where more files and materials have been created.

## (Slightly More) Advanced Git

For example, after working for a while on the project, you could have added two files inside well-organized subfolders of the project. The project would like this when viewed with a file explorer:

```bash
git-example/
|-- admin/
    |-- ethics-info.pdf
|-- experiment-1/
    |-- analysis/
        |-- plan_n.R
|-- manuscript/
|-- README
```

Running `git status` now tells that there are two new files (possible empty folders are ignored). That is, since the previous commit, two files have been added to the project; one a pdf with some administrative information (`ethics-info.pdf`), the other one is an R [@r_core_team_r:_2017] script for a pre-planned power analysis (`plan_n.R`). You would probably like to track any changes to the power analysis script, but the ethics information file isn't really something that you need to keep track of---at least for this example. It would be laborious to constantly keep ignoring it when committing changes to Git (especially if we'd like to commit multiple files simultaneously with `git add .`) Git has an elegant solution to specifying which files to keep track of. Because by default all files are monitored, Git uses a special file for instructing which files are to be *ignored*.

@@

To make Git ignore files, you simply add a plain text file called `.gitignore` to the home folder of the repository. You can use any text editor to create this file. Each row of this file should specify a file or a folder (or a regular expression) that Git should ignore. In the current example you could make Git ignore the `admin/` folder entirely, and any file with the .pdf extension inside `manuscript/`. The example .gitignore file would look like this (lines beginning with `#` are comments):

@@

# Ignore all .pdf files in the 'manuscript' folder
manuscript/*.pdf

# Ignore a specific file in a specific folder
analysis/sandbox.R
```

@@

Because there are now two untracked files that are not in .gitignore (`.gitignore` and `plan_n.R`), and you usually should aim to maintain a clean history for the project, you can create two separate commits: One for the Git ignore file, and one for the power analysis file.

@@

After this last commit, you can at any time come back to this commit with `git log` or `git show` and see what was inside the newly created power analysis file when it was first created. For instance, if new information suggests that you should change the assumed effect size in the power analysis, you can simply edit and save the file, then add and commit the changes to Git with a helpful message that logs this important event in Git's history.

This possibility of "rewinding history" is especially useful for files that might go under multiple revisions (manuscripts, analysis files), or if you are interested when and in what order the files were created---and who created or changed them. One might consider committing a power analysis as a small personal pre-registration of a part of the research plan.

@@

[^git-version]: As of the writing of this article, the current version of Git is 2.13.1.
[^git-name]: The creator of Git, Linus Torvalds, named Git after himself as "the stupid content tracker" (@mcmillan_after_2005; <https://git-scm.com/docs/git.html>)
[^simple]: By "simple", we only mean that the technical aspects of the project, related to curating materials related to it, are simple. We of course do not suggest that there might be anything theoretically or scientifically simple in such projects.
[^vcs-collaborate]: The software we present below is used used by major software developers such as Microsoft, Google, and Facebook on code bases with hundreds of contributors.
[^cli-links]: If using a text-based command line seems challenging, Codecademy (<https://www.codecademy.com/learn/learn-the-command-line>) has a free interactive online tutorial, and MIT offers a free online game to teach using the command line (<http://web.mit.edu/mprat/Public/web/Terminus/Web/main.html>).
[^git-link-submodules]: For advanced users, Git *submodules* allow linking projects to each other, or organizing more complex projects into projects and their sub-projects (<https://git-scm.com/book/en/v2/Git-Tools-Submodules>).
[^git-cheat-sheet]: It helps to have a Git command cheat sheet (<https://services.github.com/resources/cheatsheets/>) printed and taped on your wall, but it contains many more commands than are needed for the basic use of Git.
[^tier-link]: <http://www.projecttier.org/tier-protocol/specifications/>
